---
title: "Nairobi UberV3"
author: "Charles Hatfield & Marcel Reinmuth"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message=FALSE, warning=FALSE)
```

This is the exploratory analysis for school in session versus break and the impacts on mean road speeds. This purely focused on the change in speeds by road segment. Routing & traveltime will be investigated elsewhere.

# Setup

## Libraries

The following libraries are used for in this R markdown document.

```{r "Libraries", warning = FALSE, message = FALSE}
library(sf)
library(tidyverse)
library(rgeoboundaries)
library(osmextract)
library(tmap)
library(kableExtra)
library(lubridate)
library(DT)

```

## Data acquisition & preprocess

We use free and open speed traffic data from Uber for Nairobi, available here: [https://movement.uber.com/cities/nairobi/downloads/speeds?lang=en-US&tp[y]=2019&tp[q]=1](https://movement.uber.com/cities/nairobi/downloads/speeds?lang=en-US&tp%5By%5D=2019&tp%5Bq%5D=1){.uri}

Uber provides a toolkit software via npm to generate the respective road segments in a geo format. The package is available here: <https://www.npmjs.com/package/movement-data-toolkit>. We generated the road segments as geojson.

**Important**: Uber movement data covers the period 2018-2020. Via the attribute `osm_way_id` the speed traffic information can be linked to the OpenStreetMap (OSM) road network. It is not clear how well the ids from the covered period match with the current OSM data. OSM ids are stable when attributes or the geometry of a object is changed. Howevery newly added data and deleted objects are potentially ommitted when using uber traffic information from years ago but current OSM road network data.

```{r "Data Loading", warning=FALSE, message=FALSE}
#uber <- read_csv("movement-speeds-quarterly-by-hod-nairobi-2018-Q1.csv")
#uber_jan <- read_csv("movement-speeds-hourly-nairobi-2019-1.csv")
#uber_april <- read_csv("movement-speeds-hourly-nairobi-2019-4.csv")
nairobi_roads <- st_read("nairobi_2019.geojson")
st_crs(nairobi_roads) <- 4326

```

We are aggregating and averaging road segment speeds across the semester and holiday periods


```{r "Set School and Holiday dfs"}
path <- "movement-speed/"

speed_files <- list.files(path=path, pattern="movement-speeds-hourly-nairobi-2019-.*\\.csv", recursive=TRUE, full.names=TRUE)

# Read the files into a list of data frames
df_list <- lapply(speed_files, read.csv)

# Combine the data frames in the list into one data frame
year_df <- do.call(rbind, df_list)

```

```{r}
#saveRDS(year_df, "year_df.rds")
#year_df <- readRDS("year_df.rds")
```


```{r}
whole_year <- data.frame(
  date = seq(as.Date("2019-01-01"), as.Date("2019-12-31"), by = "day")
) |>
  mutate(
    weekday = wday(date, label = TRUE),
    day = day(date),
    month = month(date)
  ) |>
  filter(
    !(weekday %in% c("Sat", "Sun")) &
    !(date >= as.Date("2019-02-18") & date <= as.Date("2019-02-24")) &
    !(date >= as.Date("2019-06-17") & date <= as.Date("2019-06-23")) &
    !(date >= as.Date("2019-11-04") & date <= as.Date("2019-11-29")) # remove midterm breaks and exam period
  )

# Define the school semesters
school_periods <- whole_year |>
  mutate(
    status = case_when(
      (date >= as.Date("2019-01-02") & date <= as.Date("2019-04-05")) ~ "semester1",
      (date >  as.Date("2019-04-05") & date <  as.Date("2019-04-29")) ~ "holiday1",
      (date >= as.Date("2019-04-29") & date <= as.Date("2019-08-02")) ~ "semester2",
      (date >  as.Date("2019-08-02") & date <  as.Date("2019-08-26")) ~ "holiday2",
      (date >= as.Date("2019-08-26") & date <= as.Date("2019-10-25")) ~ "semester3",
      (date >  as.Date("2019-10-25")) ~ "holiday3",
      TRUE ~ "other" # this is just a catch-all; remove if not needed
    )
  )

periods <- split(school_periods, school_periods$status)

```


## Filtering and processing

```{r}

year_avg <- year_df |>
  group_by(hour, segment_id, start_junction_id, end_junction_id, 
           osm_way_id, osm_start_node_id, osm_end_node_id)  |>  
  dplyr::summarise(mean_speed_kph = mean(speed_kph_mean, na.rm = TRUE), .groups = "drop")

```


```{r}
filter_and_aggregate <- function(df, date_df) {
  df |>
    filter(day %in% date_df$day & month %in% date_df$month) |> 
    group_by(hour, segment_id, start_junction_id, end_junction_id, 
             osm_way_id, osm_start_node_id, osm_end_node_id)  |>  
    dplyr::summarise(mean_speed_kph = mean(speed_kph_mean, na.rm = TRUE), .groups = "drop")
}

# Applying filter_and_aggregate for each period
results <- lapply(periods, filter_and_aggregate, df = year_df)

# Adding period names to each dataframe in results
results <- mapply(function(df, period_name) {
  df$Period <- period_name
  return(df)
}, results, names(results), SIMPLIFY = FALSE)

# Convert results to separate dataframes

list2env(results, envir = .GlobalEnv) ## just use results and don't use list2env above
```

```{r}
rm(whole_year, df_list, periods, year_df, school_periods, other)
```

```{r}
## just use results and don't use list2env above
join_fields <- c("hour","start_junction_id" , "end_junction_id", 
                 "osm_start_node_id", "osm_end_node_id", 
                 "segment_id", "osm_way_id")

# Combine all your dataframes in a list
all_dfs <- list(semester1, semester2, semester3, holiday1, holiday2, holiday3, year_avg)

# Perform iterative inner_join
matched_roads <- Reduce(function(x, y) inner_join(x, y, by = join_fields), all_dfs)
matched_roads <- matched_roads |> 
  select(join_fields)


# Extract each dataframe from the list
semester1 <- inner_join(semester1, matched_roads)
semester2 <- inner_join(semester2, matched_roads)
semester3 <- inner_join(semester3, matched_roads)
holiday1  <- inner_join(holiday1, matched_roads)
holiday2  <- inner_join(holiday2, matched_roads)
holiday3  <- inner_join(holiday3, matched_roads)
year_avg  <- inner_join(year_avg, matched_roads)
```

```{r}
rm(all_dfs, other, matched_roads)
```

## Hourly boxplots

```{r}
combined_datasets <- list(
  combined_school = rbind(semester1, semester2, semester3),
  combined_holiday = rbind(holiday1, holiday2),
  combined_period1 = rbind(semester1, holiday1),
  combined_period2 = rbind(semester2, holiday2),
  combined_period3 = rbind(semester3, holiday3),
  combined_semesters = rbind(semester1, semester2, semester3),
  combined_holidays = rbind(holiday1, holiday2)
)

# All combined datasets into one dataframe
combined_average <- do.call(rbind, combined_datasets)

# Create a plotting function
plot_comparison <- function(data, periods) {
  # Assign colors dynamically based on periods
  color_values <- c("firebrick1", "dodgerblue3", "forestgreen")[seq_along(periods)]
  colors <- setNames(color_values, periods)
  
  data |> 
    filter(Period %in% periods) |> 
    ggplot(aes(x = factor(hour), y = mean_speed_kph, fill = Period)) +
    geom_boxplot(alpha = 0.5, outlier.shape = NA) +
    theme_classic() +
    scale_fill_manual(values = colors) + 
    labs(x = "Hour of the Day",
         y = "Mean Speed kph",
         fill = "Period")
}

# Render plots iteratively
plots <- list(
  plot1 = plot_comparison(combined_average, c("semester1", "holiday1")),
  plot2 = plot_comparison(combined_average, c("semester2", "holiday2")),
  plot3 = plot_comparison(combined_average, c("semester3", "holiday3")),
  plot4 = plot_comparison(combined_average, c("semester1", "semester2", "semester3")),
  plot5 = plot_comparison(combined_average, c("holiday1", "holiday2", "holiday3")),
  plot6 = plot_comparison(combined_average, c("combined_school", "combined_holiday"))
)

```

```{r}
# Print each plot
lapply(plots, print)

# Clear memory
rm(plots)
gc()
```

## Morning and afternoon splits

```{r}
# Define a function to obtain morn and afternoon periods
get_time_periods <- function(df) {
  list(
    morn = df |> 
      filter(hour %in% c(6,7,8)) |> 
      group_by(segment_id, start_junction_id, end_junction_id, 
               osm_way_id, osm_start_node_id, osm_end_node_id) |>  
      dplyr::summarise(mean_speed_kph = mean(mean_speed_kph, na.rm = TRUE), .groups = "drop"),
    
    afternoon = df |> 
      filter(hour %in% c(15,16,17)) |> 
      group_by(segment_id, start_junction_id, end_junction_id, 
               osm_way_id, osm_start_node_id, osm_end_node_id) |>  
      dplyr::summarise(mean_speed_kph = mean(mean_speed_kph, na.rm = TRUE), .groups = "drop")
  )
}

# Apply the function on all relevant dataframes
combined_school_split <- get_time_periods(combined_datasets$combined_school)
combined_holiday_split <- get_time_periods(combined_datasets$combined_holiday)

semester1_split <- get_time_periods(semester1)
semester2_split <- get_time_periods(semester2)
semester3_split <- get_time_periods(semester3)

holiday1_split <- get_time_periods(holiday1)
holiday2_split <- get_time_periods(holiday2)
holiday3_split <- get_time_periods(holiday3)

year_split <- get_time_periods(year_avg)
```

```{r}
rm(semester1, semester2, semester3, holiday1, holiday2, holiday3, combined_average, combined_datasets)
```

## Speed difference calcs

```{r}
join_fields <- join_fields[-which(join_fields == "hour")]


calculate_diff <- function(df1, df2, new_col_name) {
  # Ensure that the provided data frames are indeed data frames
  df1 <- as.data.frame(df1)
  df2 <- as.data.frame(df2)

  if(is.null(df1) || is.null(df2)) {
    stop("One of the data frames provided to 'calculate_diff' is NULL.")
  }

  # Compute the difference and clean the resulting data frame
  result <- df1 |>
    left_join(df2, by = join_fields, suffix = c(".x", ".y")) |>
    mutate(!!new_col_name := `mean_speed_kph.y` - `mean_speed_kph.x`) |>
    filter(!is.na(!!sym(new_col_name))) 

  return(result)
}

```


```{r}
## semester 1
semester1_split$morn <- semester1_split$morn |> 
  calculate_diff(year_split$morn, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

semester1_split$morn <- calculate_diff(semester1_split$morn, holiday1_split$morn, "kph_diff_hol")|> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

semester1_split$morn <- calculate_diff(semester1_split$morn, combined_holiday_split$morn, "kph_diff_fhol")|> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

semester1_split$afternoon <- calculate_diff(semester1_split$afternoon, year_split$afternoon, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester1_split$afternoon <- calculate_diff(semester1_split$afternoon, holiday1_split$afternoon, "kph_diff_hol")|> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester1_split$afternoon <- calculate_diff(semester1_split$afternoon, combined_holiday_split$afternoon, "kph_diff_fhol")|> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)


## semester 2
semester2_split$morn <- calculate_diff(semester2_split$morn, year_split$morn, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester2_split$morn <- calculate_diff(semester2_split$morn, holiday2_split$morn, "kph_diff_hol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester2_split$morn <- calculate_diff(semester2_split$morn, combined_holiday_split$morn, "kph_diff_fhol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

semester2_split$afternoon <- calculate_diff(semester2_split$afternoon, year_split$afternoon, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester2_split$afternoon <- calculate_diff(semester2_split$afternoon, holiday2_split$afternoon, "kph_diff_hol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester2_split$afternoon <- calculate_diff(semester2_split$afternoon, combined_holiday_split$afternoon, "kph_diff_fhol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

## semester 3
semester3_split$morn <- calculate_diff(semester3_split$morn, year_split$morn, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester3_split$morn <- calculate_diff(semester3_split$morn, holiday3_split$morn, "kph_diff_hol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester3_split$morn <- calculate_diff(semester3_split$morn, combined_holiday_split$morn, "kph_diff_fhol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

semester3_split$afternoon <- calculate_diff(semester3_split$afternoon, year_split$afternoon, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester3_split$afternoon <- calculate_diff(semester3_split$afternoon, holiday3_split$afternoon, "kph_diff_hol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
semester3_split$afternoon <- calculate_diff(semester3_split$afternoon, combined_holiday_split$afternoon, "kph_diff_fhol") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)


## holdays
holiday1_split$morn <- calculate_diff(holiday1_split$morn, year_split$morn, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
holiday2_split$morn <- calculate_diff(holiday2_split$morn, year_split$morn, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
holiday3_split$morn <- calculate_diff(holiday3_split$morn, year_split$morn, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

holiday1_split$afternoon <- calculate_diff(holiday1_split$afternoon, year_split$afternoon, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
holiday2_split$afternoon <- calculate_diff(holiday2_split$afternoon, year_split$afternoon, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)
holiday3_split$afternoon <- calculate_diff(holiday3_split$afternoon, year_split$afternoon, "kph_diff_year") |> 
  select(-c(mean_speed_kph.y)) |> 
  rename(mean_speed_kph = mean_speed_kph.x)

```

## Spatial join to Nairobi roads


```{r}
semester3_split$afternoon <- semester3_split$afternoon |> 
        left_join(nairobi_roads, by = c("osm_start_node_id" = "osmstartnodeid", "osm_end_node_id" = "osmendnodeid")) |> 
  st_as_sf()

st_write(semester3_split$afternoon, "semester3_afternoon.gpkg")

```

## ors output

```{r}
ors_df <- function(df, quarter) {
  df <- df  |>  
    mutate(year = 2019,
           quarter = quarter,
           speed_kph_stddev = 1,
           speed_kph_mean = 1)  |>   
    rename(speed_kph_p50 = mean_speed_kph) |> 
    select(-c("segment_id", "start_junction_id", "end_junction_id", 
              "kph_diff_hol", "kph_diff_fhol", "kph_diff_year")) |> 
    slice(rep(1:n(), each = 3)) |> 
    mutate(hour_of_day = rep(6:8, length.out = n())) |> 
    select(c("year", "quarter", "hour_of_day", "osm_way_id", 
             "osm_start_node_id", "osm_end_node_id", 
             "speed_kph_mean", "speed_kph_stddev", "speed_kph_p50"))
  return(df)
}


ors_holdf <- function(df, quarter) {
  df <- df  |>  
    mutate(year = 2019,
           quarter = quarter,
           speed_kph_stddev = 1,
           speed_kph_mean = 1)  |>   
    rename(speed_kph_p50 = mean_speed_kph) |> 
    select(-c("segment_id", "start_junction_id", "end_junction_id", 
              "kph_diff_year")) |> 
    slice(rep(1:n(), each = 3)) |> 
    mutate(hour_of_day = rep(6:8, length.out = n())) |> 
    select(c("year", "quarter", "hour_of_day", "osm_way_id", 
             "osm_start_node_id", "osm_end_node_id", 
             "speed_kph_mean", "speed_kph_stddev", "speed_kph_p50"))
  return(df)
}
```

```{r}
df_list <- list(
  "semester1_morn_ors" = ors_df(semester1_split$morn, 1),
  "semester1_afternoon_ors" = ors_df(semester1_split$afternoon, 1),
  
  "semester2_morn_ors" = ors_df(semester2_split$morn, 2),
  "semester2_afternoon_ors" = ors_df(semester2_split$afternoon, 2),
  
  "semester3_morn_ors" = ors_df(semester3_split$morn, 3),
  "semester3_afternoon_ors" = ors_df(semester3_split$afternoon, 3),

  "holiday1_morn_ors" = ors_holdf(holiday1_split$morn, 1),
  "holiday1_afternoon_ors" = ors_holdf(holiday1_split$afternoon, 1),
  
  "holiday2_morn_ors" = ors_holdf(holiday2_split$morn, 2),
  "holiday2_afternoon_ors" = ors_holdf(holiday2_split$afternoon, 2),

  "holiday3_morn_ors" = ors_holdf(holiday3_split$morn, 4),
  "holiday3_afternoon_ors" = ors_holdf(holiday3_split$afternoon, 4)
)

# Loop through the list and write each dataframe to a CSV
for (name in names(df_list)) {
  write.csv(df_list[[name]], paste0(name, ".csv"), row.names = FALSE)
}


```


```{r}
write_csv()
```

