---
title: "Nairobi School data"
author: "Marcel Reinmuth"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
    theme: cerulean
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(osmdata)
library(tmap)
library(exactextractr)
library(terra)

knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message=FALSE, warning=FALSE)
```

This document gives insights on school location data available for Nairobi, Kenya.

We look at data from WorldBank and OpenStreetMap (via Overpass API).
For OpenStreetMap a little processing is necessary due to its datamodel and possibility of dual representation of schools as POI and building footprint. In the following chunk we download the data from OSM and check for overlaps of footprints and POIs. 

```{r Data load & preparation, echo=FALSE}
nairobi_boundary <- st_read("data/nairobi_boundary.gpkg", quiet=T)

osm_schools <- opq(bbox = st_bbox(nairobi_boundary) |> as.numeric()) |>
  add_osm_feature(key = 'amenity', value = 'school') |>
  osmdata_sf()

osm_schools_poi <- osm_schools$osm_points |>
  filter(amenity=="school")

osm_schools_poly <- osm_schools$osm_polygons |>
  filter(amenity=="school")

osm_schools_multipoly <- osm_schools$osm_multipolygons |>
  filter(amenity=="school")


# filter out poly overlaps
osm_schools_poi <- st_filter(osm_schools_poi, osm_schools_poly |> st_union(), .predicate = st_disjoint)
# filter out multi poly overlaps
osm_schools_poi <- st_filter(osm_schools_poi, osm_schools_multipoly |> st_union(), .predicate = st_disjoint)

# filter out poly overlaps
osm_schools_poly <- st_filter(osm_schools_poly, osm_schools_multipoly |> st_union(), .predicate = st_disjoint)

# create point of surfaces

osm_schools <- bind_rows(
  osm_schools_poi,
  osm_schools_poly |> st_point_on_surface(),
  osm_schools_multipoly |> st_point_on_surface()
)

# read wordlbank school data https://datacatalog.worldbank.org/search/dataset/0038039
schools_wb <- st_read("data/schools/Schools/Schools.shp", quiet=T)
hex_grid <- readRDS("hex_grid.rds")
wpop <- rast("data/ken_ppp_2020_UNadj_constrained.tif")


hex_grid$count_wb <- st_intersects(hex_grid, schools_wb |> st_transform(4326)) |> lengths()
hex_grid$count_osm <- st_intersects(hex_grid, osm_schools) |> lengths()
hex_grid$count_diff_wb_osm <- hex_grid$count_wb - hex_grid$count_osm
hex_grid$wpop <- exact_extract(wpop, hex_grid, "sum", progress=F)
```


```{r Numbers}
osm_schools |> nrow()
selected_cols <- grep("education|name|osm_id|amenity", names(osm_schools), ignore.case = TRUE)
osm_schools <- osm_schools |> select(all_of(selected_cols))

osm_schools <- osm_schools |>  st_filter(nairobi_boundary)


schools_wb <- schools_wb |>
  st_transform(4326) |>
  st_filter(nairobi_boundary)

schools_wb |> names()

# How many schools overlap within different distances
wb50_osm <- osm_schools |> st_transform(32637) |>
  st_filter(schools_wb |>
              st_transform(32637) |>
              st_buffer(50) |> st_union()) |> nrow()

wb100_osm <- osm_schools |> st_transform(32637) |>
  st_filter(schools_wb |>
              st_transform(32637) |>
              st_buffer(100) |> st_union()) |> nrow()


wb500_osm <- osm_schools |> st_transform(32637) |>
  st_filter(schools_wb |>
              st_transform(32637) |>
              st_buffer(500) |> st_union()) |> nrow()


wb1000_osm  <- osm_schools |> st_transform(32637) |>
  st_filter(schools_wb |>
              st_transform(32637) |>
              st_buffer(1000) |> st_union()) |> nrow()

osm50_wb <- schools_wb |> st_transform(32637) |>
  st_filter(osm_schools |>
              st_transform(32637) |>
              st_buffer(50) |> st_union()) |> nrow()

osm100_wb <- schools_wb |> st_transform(32637) |>
  st_filter(osm_schools |>
              st_transform(32637) |>
              st_buffer(100) |> st_union()) |> nrow()


osm500_wb <- schools_wb |> st_transform(32637) |>
  st_filter(osm_schools |>
              st_transform(32637) |>
              st_buffer(500) |> st_union()) |> nrow()


osm1000_wb  <- schools_wb |> st_transform(32637) |>
  st_filter(osm_schools |>
              st_transform(32637) |>
              st_buffer(1000) |> st_union()) |> nrow()


```

# Overlap of schools in different datasets

Total amount of schools per dataset.

| Dataset | Total amount of schools |
|---|---|---|
| OSM | `r osm_schools |> nrow()` |
| Worldbank | `r schools_wb |> nrow()` |


Overlap by distance from Wordlbank


| Coverage of WB  | 50m | % | 100m | % | 500m | % | 1000m | % |
|---|---|---|---|---|---|---|---|---|
| OSM | `r wb50_osm` | `r (wb50_osm / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` | `r wb100_osm` | `r (wb100_osm / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` | `r wb500_osm ` | `r (wb500_osm / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` | `r wb1000_osm` | `r (wb1000_osm / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` |


Overlap by distance from OpenStreetMap


| Coverage of OSM  | 50m | % | 100m | % | 500m | % | 1000m | % |
|---|---|---|---|---|---|---|---|---|
| WB | `r osm50_wb` | `r (osm50_wb / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` | `r osm100_wb` | `r (osm100_wb / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` | `r osm500_wb` | `r (osm500_wb / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` | `r osm1000_wb` | `r (osm1000_wb / osm_schools |> nrow() * 100 |> as.numeric()) |> round(2)` |


# Comparing locations of schools

```{r Maps 1}
tmap_mode("view")

m.1 <- tm_basemap("OpenStreetMap") +
  tm_shape(osm_schools) +
  tm_dots(
    col="turquoise",
    size = .1,
    border.lwd = 0,
    popup.vars = c("osm_id", "amenity", "name", "education:students")
  )

m.2 <- tm_basemap("OpenStreetMap") +
  tm_shape(schools_wb) +
  tm_dots(
    col="firebrick",
    size = .1,
    border.lwd = 0,
    popup.vars = c("SCHOOL_NAM", "LEVEL", "Status", "Source")
  )

tmap_arrange(m.1,m.2, sync = T)


```

# Aggregated on Hex grid

Also added a difference and worldpop map.

```{r Maps 2}
colrmp <- viridisLite::cividis(7)
colrmp[1] <- "#FFFFFF"



tmap_mode("view")

m.1 <- tm_basemap("OpenStreetMap") +
  tm_shape(hex_grid) +
  tm_polygons(
    title = "Schools in WB",
    col="count_wb",
    border.lwd = .1,
    breaks=c(0,1, 10, 20, 30, 50, 100, 150),
    palette=colrmp
  )

m.2 <- tm_basemap("OpenStreetMap") +
  tm_shape(hex_grid) +
  tm_polygons(
    title = "Schools in OSM",
    col="count_osm",
    border.lwd = .1,
    breaks=c(0,1, 10, 20, 30, 50, 100, 150),
    palette=colrmp
  )

m.3 <- tm_basemap("OpenStreetMap") +
  tm_shape(hex_grid) +
  tm_polygons(
    title = "WB - OSM counts",
    col="count_diff_wb_osm",
    breaks=c(-100,-50,-10,-1,0,1,10,50,100),
    border.lwd = .1,
    palette="RdBu"
  )

m.4 <- tm_basemap("OpenStreetMap") +
  tm_shape(hex_grid) +
  tm_polygons(
    title = "Population WorldPop",
    col="wpop",
    border.lwd = .1,
    breaks=c(0, 1, 100, 1000, 10000, 50000, 150000),
    palette=colrmp
  )

tmap_arrange(m.1,m.2,m.3,m.4,ncol = 2, nrow = 2, sync = T)


```
